---
title: "Outline (Inclass notes)"
output: html_document
date: "2023-09-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

### 0. Load the `tidyverse` Packages

This section loads the packages we need in this lecture.

```{r}
library(tidyverse)
```

### 1. Import and Export the V-Dem Data

This section loads the VDEM dataset and describe its basic information

```{r}
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

### 2. to take a subset of *columns* in the existing data: `select`

Our second step is to select economic development indicators

First we look at the identifiers of the indicator

```{r}
names(d)
```

Select country year column --\> only distinct

Select country year column --\> only distinct country_names

```{r}
d |> select(country_name, country_id, year) |>
  distinct(country_name)
```

Select only e_gdp (GDP) and e_gdppc (GDP per capita)

```{r}
d |> select(country_name, country_id, year, e_gdp, e_gdppc)
```

Create a new dataset

```{r}
d_gdp <- d |>
  select(country_name, country_id, year, e_gdp, e_gdppc)
```

### 3. to rename columns: `rename`

```{r}
d_gdp <- d_gdp |>
  rename("GDP" = "e_gdp","GDP_per_Capita" = "e_gdppc", "Country" = "country_name", "ID" = "country_id", "Year" = "year")
```

### 4. to take a subset of *rows* by some simple conditions: `slice_`

get countries-years with highest GDP

d

```{r}
d_gdp |>
  slice_max(order_by = GDP, n = 10)
```

get random sample (every time you run, you get a different result)

```{r}
d_gdp |>
  slice_sample(n = 10)

```

```{r}
# Want countries-years with highest GDP
d_gdp |> slice_max(order_by = GDP, n = 10)

## # A tibble: 10 × 5
##    Country                     ID  Year      GDP GDP_per_capita
##    <chr>                    <dbl> <dbl>    <dbl>          <dbl>
##  1 China                      110  2019 2279809.           15.4
##  2 China                      110  2018 2205730.           14.9
##  3 China                      110  2017 2136176.           14.5
##  4 United States of America    20  2019 2118706.           60.6
##  5 United States of America    20  2018 2077898.           59.6
##  6 China                      110  2016 2039529.           13.9
##  7 United States of America    20  2017 2023242.           58.5
##  8 United States of America    20  2016 1980809.           57.6
##  9 China                      110  2015 1953127.           13.3
## 10 United States of America    20  2015 1942092.           56.7
```

```{r}
# Get countries-years with the lowest GDP
d_gdp |> slice_min(order_by = GDP, n = 10)
```

```{r}
set.seed(52)
d_gdp |> slice_sample(n = 10) # Sample 10 observations
```

### 5. to take a subset of *rows* by some more complicated conditions: `filter`

want 2000-2005

```{r}
d_gdp |>
  filter(Year>=2000 & Year<=2005 & Country == "China")
```

```{r}
d_gdp_china <- d_gdp |> filter(Country == "China")|> filter(Country == "China")
```

```{r}
d_gdp_china
```

### 6. to sort the rows based on the value of one or multiple columns: `arrange`

arrange by GDP per Capita (asc.)

```{r}
d_gdp |>
  arrange(GDP_per_Capita)
```

arrange by GDP per Capita (desc.) use minus "-" sign

```{r}
d_gdp |>
  arrange(-GDP_per_Capita)
```

### 7. to perform (4) (5) (6) group by group: `group_by`, `ungroup`

```{r}
#want: for each country, we want the year with the highest GDP
# n=1 is mean only one record
# without groupby, it will only show one result,the higheest gdp
d_gdp |> 
  group_by(Country)|>
  slice_max(GDP,n=1)
```

```{r}
# How many entries are there for each country
d_gdp |>
  group_by(Country) |>
  count()
```

```{r}
# to look up the fouction of count
?count
```

```{r}
# want: for each Year, get the year when countries has the worst GDP
d_gdp |>
  group_by(Country) |>
  slice_min(order_by = GDP, n=1)
```

### 8. to create new columns in the data: `group_by`, `mutate`, `ungroup`

```{r}
#create a new colum and take the value "1"
d_gdp |>
  mutate(New = 1)

d_gdp |>
  mutate(New=GDP)

d_gdp |> mutate(New=log(GDP))

d_gdp |>
  mutate(New=log(GDP)+1)

# Want: New colum to be GDp relative to average GDP in the world 1984-2022
d_gdp |>mutate(GDP_over_avg =GDP / mean(GDP, na.rm=TRUE))

# Want: New column to be gdp relaive to average  gdp of the country in the world 1984-2022
d_gdp |>
  group_by(Country) |>
  mutate(GDP_over_avg =GDP / mean(GDP, na.rm=TRUE))
```

```{r}
#Country-year development level with referene to that of 1984
?first
```

```{r}
d_gdp |>
  group_by(Country) |>
  arrange(Year) |>
  mutate(GDP_over_1984 = GDP/first(GDP)) |>
  ungroup()|>
  arrange(Country,Year)
```

```{r}
# Year-on-year economic growth
#?lag

d_gdp |>
  group_by(Country) |>
  arrange(Year) |>
  mutate(GDP_yoy_change = GDP- lag(GDP, n=1)) |>
  ungroup() |>
  arrange(Country,Year)
```

### 9. to summarize the data: `group_by`, `summarise`, `ungroup`

```{r}
# Want: Average GDP level of the world
d_gdp |> summarise(gdp_average = mean(GDP, na.rm= TRUE),
                   GDP_per_Capita= mean(GDP_per_Capita, na.rm = TRUE))
```

```{r}
# Average develoment level
d_gdp |>
  group_by(Country)|>
  summarise(GDP_average = mean(GDP, na.rm= TRUE), GDP_per_Capita = mean(GDP_per_Capita, na.rm=TRUE))
```

```{r}
# GDP growth and  GDP per capita growth: 2020 over 1984
d_gdp |>
  filter(Year >= 1984, Year <= 2020) |>
  group_by(Country) |>
  arrange(Year) |>
  summarise(GDP_growth_2020_1984 = (last(GDP)-first(GDP))/first(GDP)) |>
  ungroup()|>
  arrange(Country)
```

```{r}
# Data integrity check
d_gdp |>
  group_by(Country) |>
  # Create a column that indicates whether the value is missing
  mutate(GDP_missing = as.numeric(is.na(GDP)), .after = GDP) |>
  group_by(Country) |>
  summarise(N_GDP_missing = sum(GDP_missing))
```

#### 

# Lecture 5

### 1. Reshape the V-Dem dataset

```{r}
d |>select(country_name) |> distinct()

d_subset <- d |>select(country_name, year, starts_with("e_"))


d_subset_long <- 
  d_subset |> pivot_longer(cols = starts_with("e_"))
```

### 2. Stack multiple subsets of the V-Dem datasets

```{r}
d_VemIndex <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

```{r}
d_stack <- bind_cols(d,d_VemIndex)
```

```{r}
# Want: stack two tables by rows?
d_1945_1983 <-  read_csv("_DataPublic_/vdem/1945_1983/vdem_1945_1983_external.csv")
```

```{r}
d_1945_2022 <-bind_rows(d, d_1945_1983)
  
d_1945_2022 |> select(year)|> distinct() |> arrange(year)
```
