---
title: "Lecture9_note_areal"
author: "Shukyee Chan"
date: "2023-11-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 9, fig.height = 3, message = FALSE, warning = FALSE)
```

## Load the Packages

```{r}
library(tidyverse)
library(ggthemes)
theme_set(theme_map())
```

## Load the Conflict Data

```{r}
d = read_csv("Lecture_09/data/GEDEvent_v22_1.csv")
names(d)
```

## Map Inter-Country Variation

Draw a world map featuring the variation of conflicts among countries in the year of 2021.

### Put together your data

```{r}
# Create a dataframe named d_country_2021 that summarizes the number of conflicts in each country in the year of 2021.
d_country_2021 = d |>
  filter(year == 2021) |>
  group_by(country, country_id, region) |>
  summarise(
    n_conflict = n()
  )

d_country_2021 = d_country_2021 |> arrange(-n_conflict)
```

## Load the World Map

```{r}
world = map_data("world")

names(world)

# Summary statistics
summary(world)

# Problem: Max. longitude > 180! Remove those
world = world |> filter(long <= 180)
```

```{r}
# Plot the world map
world |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region),
    color = "white", fill = "gray", size = 0.1
  ) +
  coord_map()
```

## Merge Data with the World Map

```{r}
world_m = world |>
  left_join(d_country_2021, by = c("region" = "country"))
```

## Visualize Conflict Data on a Map

```{r}
# Color the countries with the number of conflicts in 2021.
world_m |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = n_conflict),
    color = "black", size = 0.1
  ) +
  coord_map()
```

## Make the Map More Informative

### Change the color scheme `scale_fill_viridis_c()`

```{r}
world_m |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = n_conflict),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "A") +
  coord_map()

world_m |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = n_conflict),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "B") +
  coord_map()

world_m |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = n_conflict),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "B", direction = -1) +
  coord_map()

world_m |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = n_conflict),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "D", direction = -1) +
  coord_map()
```

### Change the legend's name

```{r}
# The most robust way to change the legend's name is the rename the variable.
# When your variable name contains space, you should put "`" around it.
world_m |>
  rename("Number of Conflicts" = "n_conflict") |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = `Number of Conflicts`),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "B", direction = -1) +
  coord_map()
```

### Transforming the Scale or Data

```{r}
# The distribution of the conflict frequencies is so skewed that makes the differences of low-conflict countries indistinguishable. Transform the data or the scales to take into account the skewed data.
world_m |>
  rename("Number of Conflicts" = "n_conflict") |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = `Number of Conflicts`),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "B", direction = -1, trans = "log") +
  coord_map()
```

Or transform the data

```{r}
world_m |>
  rename("Number of Conflicts" = "n_conflict") |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = log(`Number of Conflicts`+1)),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "B", direction = -1) +
  coord_map() +
  theme(legend.position = "bottom")
```
