---
title: "lecture5_note_data_wrangling"
author: "Shukyee Chan"
date: "2023-10-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

1.  ... to *reshape* a table (long \<-\> wide) with `pivot_longer` and `pivot_wider`
2.  ... to *stack* tables by row or by column with `bind_rows` and `bind_cols` (or, alternatively, `cbind` and `rbind`)
3.  ... to *merge* two tables with `inner_join`, `full_join`, `left_join`, `right_join`, `semi_join`, and `anti_join`

## Outline of In-Class Demo

For this in-class demonstration, we will continue working on the external parts of the V-Dem data from 1984 to 2022. The data are located here: `_DataPublic_/vdem/1984_2022/vdem_1984_2022_external`

1.  Reshape the V-Dem dataset
    1.  `pivot_longer`: Make it a long table where each variable gets its own row. That is, a row in the new dataset is a *country-year-observation*.
    2.  `pivot_wider`: Widen the above long table so that each *Year* has its own column.
2.  Stack multiple subsets of the V-Dem datasets by row and by columns
    1.  `bind_cols`: Merge the following two subsets of the V-Dem data: `_DataPublic_/vdem/1984_2022/vdem_1984_2022_external` and `_DataPublic_/vdem/1984_2022/vdem_1984_2022_index`
    2.  `bind_rows`: Merge the following two subsets of the V-Dem data: `_DataPublic_/vdem/1984_2022/vdem_1984_2022_external` and `_DataPublic_/vdem/1945_1983/vdem_1945_1983_external`
3.  Join multiple regional subsets of the V-Dem datasets
    1.  Make a new data frame that contains the following variables: `country_name`, `year`, `e_regionpol_6C`, `e_fh_status`, `e_gdppc`, and `e_gdp`
    2.  Create two separate subsets of the above data frames. Each subset include a subset of countries/ regions that are within the *region* (defined by `e_regiongeo` and `e_regionpol_6C` respectively) where *China* is located.
    3.  Explore the behavior of `inner_join`, `full_join`, `left_join`, `right_join`, `semi_join`, and `anti_join` with the two data frames.
4.  Validate V-dem's GDP data with World Bank data

```{r}
library(tidyverse)
```

```{r}
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

### 1. Reshape the V-Dem dataset

```{r}
# Want: Each row contain country-year-variable 
# want to make a row contain only one variable of the country

# take a look at the names of the variable
d |> select(country_name) |> distinct()

d_subset <- d |>
  select(country_name, year,starts_with("e"))

d_subset_long <- d_subset |>
  pivot_longer(cols = starts_with("e"))

# transform: make a wide dataset with a lot of columns into each columns become has its own row
```

```{r}
d_subset_wide_year <- d_subset_long |>
  pivot_wider(names_from = year, values_from = value)
```

### 2. Stack multiple subsets of the V-Dem datasets

```{r}
d_VdemIndex <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_index.csv")
```

```{r}
d_stack <- bind_cols(d, d_VdemIndex)
```

```{r}
# Want: Stack two tables by rows
d_1945_1983 <- read_csv("_DataPublic_/vdem/1945_1983/vdem_1945_1983_external.csv")
```

```{r}
d_1945_2022 <- bind_rows(d, d_1945_1983)

d_1945_2022 |> 
  select(year) |> 
  distinct() |> 
  arrange(year)
```
