---
title: "Lecture6_note_data_wrangling"
author: "Shukyee Chan"
date: "2023-10-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Outline

-   Reshape (long \<-\> wide) with `pivot_longer` and `pivot_wider`
-   Stack tables by row or by column with `bind_rows` and `bind_cols` (or, alternatively, `cbind` and `rbind`)
-   Merge two tables with `inner_join`, `full_join`, `left_join`, `right_join`, `semi_join`, and `anti_join`
-   Save your outputs

```{r}
library(tidyverse)
```

```{r}
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
d |> print(n = 3)
```

```{r}
d_gdp <- d |> 
  select(country_text_id, year, e_gdp, e_gdppc) |>
  rename("gdp" = "e_gdp", "gdppc" = "e_gdppc")

d_gdp |> print(n = 3)
```

# 1. Reshape a Table

## Wide to Long: `pivot_longer`

```{r}
d_gdp_long <- d_gdp |>
  pivot_longer(cols = c("gdp", "gdppc"), 
               names_to = "variable", values_to = "value")

d_gdp_long |> print(n = 4)
```

## Long to Wide: `pivot_wider`

```{r}
# Reverse the above pivot_long operation
d_gdp_wide_1 <- d_gdp_long |>
  pivot_wider(names_from = "variable", values_from = "value")

d_gdp_wide_1 |> print(n = 4)
```

```{r}
# Make year the column variable
d_gdp_wide_2 <- d_gdp_long |>
  pivot_wider(names_from = "year", values_from = "value")

d_gdp_wide_2 |> print(n = 2)
```

```{r}
#  Make country_text_id the column variable
d_gdp_wide_3 <- d_gdp_long |>
  pivot_wider(names_from = "country_text_id", values_from = "value")

d_gdp_wide_3 |> print(n = 2)
```

# 2. Stack Tables

```{r}
# New data (for stack data vertically)
d_gdp_1945 <- 
  read_csv("_DataPublic_/vdem/1945_1983/vdem_1945_1983_external.csv") |>
  select(country_text_id, year, e_gdp, e_gdppc) |>
  rename("gdp" = "e_gdp", "gdppc" = "e_gdppc")

d_gdp_1906 <- 
  read_csv("_DataPublic_/vdem/1906_1944/vdem_1906_1944_external.csv") |>
  select(country_text_id, year, e_gdp, e_gdppc) |>
  rename("gdp" = "e_gdp", "gdppc" = "e_gdppc")

d_gdp_1945 |> print(n = 2) 
```

```{r}
# New data (for stack data horizontally)
d_edu <- d |>
  select(e_peaveduc, e_peedgini) |>
  rename("edu_15" = "e_peaveduc", "edu_gini" = "e_peedgini")

d_fh <- d |>
  select(starts_with("e_fh")) |>
  rename("fh_CivilLiberty" = "e_fh_cl", "fh_PoliticalRight" = "e_fh_pr",
         "fh_RuleOfLaw" = "e_fh_rol", "fh_Status" = "e_fh_status")

d_fh |> print(n = 2)
```

## `bind_rows`

```{r}
d_gdp_1945_2022 <- bind_rows(d_gdp, d_gdp_1945)
d_gdp_1945_2022 |> print(n = 3)
unique(d_gdp_1945_2022$year) |> sort()

d_gdp_1945_2022_ue_rows <- bind_rows(
 d_gdp |> select(-gdppc),
 d_gdp_1945 |> select(-gdp)
)
```

```{r}
d_gdp_1906_2022 <- bind_rows(d_gdp, d_gdp_1945, d_gdp_1906)
d_gdp_1906_2022 |> print(n = 3)
unique(d_gdp_1906_2022$year) |> sort()
```

## `bind_cols`

```{r}
d_gdp_edu_fh <- bind_cols(d_gdp, d_edu, d_fh)
d_gdp_edu_fh |> print(n = 3)
names(d_gdp_edu_fh)
```

### These are error-prone operations

-   Do `bind_rows` and `bind_cols` ONLY WHEN you know for sure that there will not be a mismatch!

-   If you have any slightest doubt, don't use them.

# 3. Join Tables

-   `left_join`: Merge and only keep observations whose identifiers (matching keys) appear in the left-hand-side table.

-   `right_join`: Merge and only keep observations whose identifiers (matching keys) appear in the right-hand-side table.

-   `inner_join`: Merge and only keep observations whose identifiers (matching keys) appear in both tables.

-   `full_join`: Merge and keep observations whose identifiers (matching keys) appear either table.

-   `anti_join`: Filter out observations whose identifiers (matching keys) appear in the right-hand-side table

-   `semi_join`: Filter out observations whose identifiers (matching keys) do not appear in the right-hand-side table

## Task 1: The Case

Join two datasets from the V-Dem data using the above different `join_` functions

-   *GDP* data from **2000-2022**

-   *GDP per capita* data from **1984 to 2010**

```{r}
# Setup 
d_gdp_2000_2022 <- d |> filter(year %in% 2000:2022) |>
  select(country_text_id, year, e_gdp) |> rename("gdp" = "e_gdp")

d_gdppc_1984_2010 <- d |> filter(year %in% 1984:2010) |>
  select(country_text_id, year, e_gdppc) |> rename("gdppc" = "e_gdppc")

d_gdp_2000_2022 |> print(n = 2)
d_gdppc_1984_2010 |> print(n = 2)
```

## `left_join`

```{r}
d_lj <- d_gdp_2000_2022 |> 
  left_join(d_gdppc_1984_2010, by = c("country_text_id", "year"))

d_lj |> print(n = 2)
unique(d_lj$year) |> sort()
```

## `right_join`

```{r}
d_rj <- d_gdp_2000_2022 |> 
  right_join(d_gdppc_1984_2010, by = c("country_text_id", "year"))

d_rj |> print(n = 2)
unique(d_rj$year) |> sort()
```

## `inner_join`

```{r}
d_ij <- d_gdp_2000_2022 |> 
  inner_join(d_gdppc_1984_2010, by = c("country_text_id", "year"))

d_ij |> print(n = 2)
unique(d_ij$year) |> sort()
```

## `full_join`

```{r}
d_fj <- d_gdp_2000_2022 |> 
  full_join(d_gdppc_1984_2010, by = c("country_text_id", "year"))

d_fj |> print(n = 2)
unique(d_fj$year) |> sort()
```

## `semi_join`

```{r}
d_sj <- d_gdp_2000_2022 |> 
  semi_join(d_gdppc_1984_2010, by = c("country_text_id", "year"))

d_sj |> print(n = 2)
unique(d_sj$year) |> sort()
```

## `anti_join`

```{r}
d_aj <- d_gdp_2000_2022 |> 
  anti_join(d_gdppc_1984_2010, by = c("country_text_id", "year"))

d_aj |> print(n = 2)
unique(d_aj$year) |> sort()
```

## 3. Join by Identifiers with Different Variable Names

two options: (1) Rename it beforehand, (2) specify the `by =` argument differently.

```{r}
# I make an artificial example whose variable name of a matching 
# identifier is different from d_gdp_2020_2022.
d_gdppc_1984_2010_t <- d_gdppc_1984_2010 |> 
  rename("country_id" = "country_text_id")
```

```{r}
# Option 1: Rename the variables beforehand
d_aj_t <- d_gdp_2000_2022 |> 
  rename("country_id" = "country_text_id") |>
  anti_join(d_gdppc_1984_2010_t, by = c("country_id", "year"))

# Option 2: Specify the "by =" argument with a *named vector*
d_aj_t_2 <- d_gdp_2000_2022 |> 
  anti_join(d_gdppc_1984_2010_t, 
            by = c("country_text_id" = "country_id", 
                   "year" = "year"))
```

## 4. Many-to-One Join: Repeat!

Calculate each country's average 1984-2010 *GDP per capita* and merge it with our annual GDP data from 2000 to 2022.

```{r}
d_gdppc_1984_2010_avg <- d_gdppc_1984_2010 |> group_by(country_text_id) |>
  summarise(gdppc_1984to2010 = mean(gdppc, na.rm = TRUE))
d_gdppc_1984_2010_avg |> print(n = 2)
```

```{r}
d_lj_ManyToOne <- d_gdp_2000_2022 |> 
  left_join(d_gdppc_1984_2010_avg, by = "country_text_id")
d_lj_ManyToOne |> print(n = 2)
```

# 5. Save Outputs

-   `.csv` "comma-separated values," readable by Excel or a text editor

-   `.rds` "R data serialization," readable by R only

```{r}
# Save to a .csv file
write_csv(d_gdp_1945_2022, "Lecture_06/data/gdp_1945_2002.csv")

# Save to a .rds file
saveRDS(d_gdp_1945_2022, "Lecture_06/data/gdp_1945_2002.rds")
```

## Saving Your Outputs after Data Wrangling

```{r}
# Read a .csv file
d_read_1 <- read_csv("Lecture_06/data/gdp_1945_2002.csv")

# Read a .rds file
d_read_2 <- readRDS("Lecture_06/data/gdp_1945_2002.rds")
```
