---
title: "Lecture4_note_data_wrangling"
author: "Shukyee Chan"
date: "2023-09-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Lecture 4: Data Wrangling

### Overview

1.  to import and export data: `read_csv` , `write_csv` (with a brief introduction to other data import/ export functions from [`readr`](https://readr.tidyverse.org/)).

2.  to take a subset of *columns* in the existing data: `select`

3.  to rename columns: `rename`

4.  to take a subset of *rows* by some simple conditions: `slice_`

5.  to take a subset of *rows* by some more complicated conditions: `filter`

6.  to sort the rows based on the value of one or multiple columns: `arrange`

7.  to perform (4) (5) (6) group by group: `group_by`, `ungroup`

8.  to create new columns in the data: `group_by`, `mutate`, `ungroup`

9.  to summarize the data: `group_by`, `summarise`, `ungroup`

### 0. load the packages

```{r}
library(tidyverse)
```

### 1. Import and Export the V-Dem

```{r}
d <- read.csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")

getwd()
```

### 2. Select economic development indicators

Look at the identifiers of the data

```{r}
# names(d)

# Alternative functions (provides information about the dataset): The str() provides not only variable names, but also their data types and a few example data points.

str(d)
```

```{r}
d |> select(country_name, country_id, year) |>
  distinct()
```

```{r}
?distinct
# "distinct" means telling it not to repeat, keep only unique/distinct rows from a data frame
```

```{r}
# Which countries are in this dataset
d |> select(country_name) |> distinct()
# if don't have the distinct function , it will show all country name including those repeat
```

```{r}
d |> select(year) |> distinct()
```

Select both the country identifier, GDP, and GDP per capita

```{r}
d_gdp <- d |> select(country_name, country_id, year, e_gdp, e_gdppc)
```

### 3. Rename columns to make names informative

```{r}
d_gdp <- d_gdp |> 
  rename("GDP" = "e_gdp", "GDP_per_capita" = "e_gdppc", "Year" = "year", "Country" = "country_name", "ID" = "country_id")

d_gdp
```

### 4. Subset Rows of the Data Using `slice_`

```{r}
# to extract a subset of rows from a dataset based on specific criteria, such as selecting rows with the largest or smallest values in a particular column.
?slice
# slice() lets you index rows by their (integer) locations. It allows you to select, remove, and duplicate rows.
```

```{r}
#Want countries-years with higest GDP
d_gdp |>
  slice_max(order_by = GDP, n = 10)
# n = 10, mean wants the top 10 counties among the whole dataset
# It just guaranteed thar giver the top 10 you specific, yet it doesn't guatantee that they are sorted properly
```

```{r}
# the case that have the lowest gdp
# Get the countries-years with the lowest GDP
d_gdp |>
  slice_min(order_by = GDP, n = 10)
```

```{r}
# what if you want some random sample
d_gdp |> 
  slice_sample(n = 10) # sample 10 observations => randomly give you 10

# sample = everytime you run it, it gives you different results

# but the report should be replicable and reproduciable (which have to get the same random results everytime_ => so need to set the random seed

set.seed(52)
d_gdp |> 
  slice_sample(n = 10)
#set.seed function is requried in every random sampling
#the seed value of 52 is chosen randomly It could have been any other integer value. The purpose of setting the seed is to obtain the same random sample of rows every time the code is executed, as long as the seed value remains the same.
```

```{r}
# if you want 10% of the data 
set.seed(52)

d_gdp |>
  slice_sample(prop = 0.1)
# 10% of the data is 678 observations

```

### 5. Subset data by row `filter`

```{r}
# Want: 2000-2005 data (conditions you want)
# new function to use is "filter"
d_gdp |> filter(Year >= 2000 & Year <= 2005)
# >= refers to less than or equal to 2000, <= refers to greater than or equal to
```

```{r}
# Want GDP data in China
d_gdp |> filter(Country == "China")
```

```{r}
# Want: 2000-2005 from china
d_gdp |> 
  filter(Year >= 2000 & Year <= 2005) |>
  filter(Country == "China")

# it works too
d_gdp |> filter(Year >= 2000 & Year <= 2005 & 
                  Country == "China")

```

### 6. Sort the Data based on Values of Rows using `arrange`

```{r}
# Want: sort the data row by GDP per capita
# lowest to largest
d_gdp |> arrange(GDP_per_capita)
```

```{r}
# largest to lowest
d_gdp |> arrange(-GDP_per_capita)
```

### 7. Perform (4)(5)(6) group by group: `grooup_by`, `ungroup`

**Task 1:** Create a dataset that, for each country, include the country-year with the highest GDP.

```{r}
# Want: for each country, we want the year with the highest GDP
# need to put this huge table into small group (each group being one data point from one county)
# for the one country, it slice the observation with the maximun
d_gdp |>
  group_by(Country) |>
  slice_max(GDP, n = 1)

# group_by itself do nothing with the dataset, it's just a function to separate the dataset into small group. So what makes a change is the thing that comes up to group by
# what comes up with group by makes the changes
```

```{r}
# how many countries are there for each country
d_gdp |> 
  group_by(Country) |>
  count()

# count function tell the numbers of rows for each group
```

```{r}
# Want: for each countries, get the year when it have the worst GDP
d_gdp |> 
  group_by(Country) |>
  slice_min(order_by = GDP, n = 1)
```

**In-class exercise:** Create a dataset that, for each country, include the country-year with the lowest GDP.

```{r}
d_gdp_lowest <- d_gdp |>
  group_by(Country) |>
  slice_min(GDP, n = 1, na_rm = TRUE)

d_gdp_lowest
```

### 8. Create new columns in the data: `group_by`, `mutate`, `ungroup`

```{r}
d_gdp |>
  mutate(New = 1)

# create a new columns, and then it takes the value, in this case the value is 1

# create a new columns that the value is exactly the same with the existing columns

d_gdp |>
  mutate(New = GDP)

# Create a new columns is the logrithem of GDP
d_gdp |>
  mutate(New = log(GDP))

# Combination of some of the existing columns -> kind of mathermatical operation
d_gdp |>
  mutate(New = log(GDP) + 1)

# create some columns that is based on some charateristics of a group
# Want: New column to be GDP relative to average GDP in the world 1984-2022
d_gdp |> 
  mutate(d_gdp_over_avg = GDP/ mean(GDP, na.rm = TRUE ))

# if without na.rm = TRUE, it wills shows NA since there are some missing value
# na.rm is to ignore the missing data

# Want: New column to be GDP relative to average GDP of the country in the world 1984-2022
# want country specific instead of the world dataset, the mean become country specofoc mean
d_gdp |>
  group_by(Country) |>
  mutate(d_gdp_over_avg = GDP/ mean(GDP, na.rm = TRUE ))

```

**Task:** Add the following economic indicators to the data:

1.  Country-year development level with reference to that of 1984.

2.  Year-on-year economic growth.

```{r}
# Country-year developmentlevel with reference to that of 1984
# which is want to knwo the GDP level ewith reference to its gdp level in 1984, that is the first year of the datasets
# so, it required to use the function first
?first
# extracting a single value from a vector,in the case of first it extract the first value of the dataset

# Want each year GDP divied by the 1984 GDP, the value of the new columns is the percentage of the year compared to the GDP of 1984
d_gdp |>
  group_by(Country) |>
  arrange(Year) |>
  mutate(GDP_over_1984 = GDP / first(GDP)) |> # it shows all the 1984 data at the start of the table, but wanna see it sorted by country and year
  ungroup() |>
  arrange(Country, Year)
```

```{r}
# Year-on-year economic development
# to see how the countries GDP is difference from the previous years, what's the growth or decline compared to the previous year
# so this year GDP minus the past year GDP
?lag
#Find the "previous" (lag()) values in a vector. Useful for comparing values behind of or ahead of the current values.
d_gdp |>
  group_by(Country) |>
  arrange(Year) |>
  mutate(GDP_yoy_change = GDP - lag(GDP, n = 1)) |>
  ungroup() |>
  arrange(Country, Year)
```

Note:

`mutate` is used for creating new columns, it is flexible in a way that that it allows to specify the relationship between the new column and the current columns

Whenever try to create some new columns that based on the existing data -\> use`mutate`, it is powerful when it couple with the `group_by` and `ungroup`

when you do `mutate`, you will still get the same dataset that have the same number of rows as the previous dataset, so the `mutate` will not change the numbers of rows.

So `mutate` only add a new columns, it add columns, but it does not change the number of rows

### 9. Summarise the data: `group_by`, `summarise`, `ungroup`

```{r}
# Want: Average GDP level of the world
d_gdp |>
  summarise(gdp_average = mean(GDP, na.rm = TRUE))

# it just summarises an existing columns and it gives a data frame of one row and one columns

d_gdp |>
  summarise(gdp_average = mean(GDP, na.rm = TRUE),
            GDP_per_capita_average = mean(GDP_per_capita, na.rm = TRUE))
```

**Task:** Perform a data availability/ integrity check. Then aggregate the data into a new country-level dataset which contains the following indicators:

1.  Average development level from 1984 to 2022.

2.  Magnitude of growth from 1984 to 2022.

```{r}
# Want: Average development level
d_gdp |>
  group_by(Country) |>
  summarise(GDP_average = mean(GDP, na.rm = TRUE),
            gdp_per_capita_average = mean(GDP_per_capita, na.rm = TRUE))

# can calculate different thins in one summarise function
```

```{r}
# GDP growth and GDP per capita growth: comparing 2019 with 1984
d_gdp |>
  filter(Year >= 1984 & Year <= 2019) |>
  group_by(Country) |>
  arrange(Year) |>
  summarise(GDP_growth_2019_1984 = (last(GDP) - first(GDP)) / first(GDP),
            GDPpc_growth_2019_1984 = (last(GDP_per_capita) - first(GDP_per_capita)) / first(GDP_per_capita)) |>
  ungroup() |>
  arrange(Country)
```

```{r}
# Data availability/ integrity check
d_gdp |>
  # Create a column that indicates whether the value is missing
  mutate(GDP_missing = as.numeric(is.na(GDP)), .after = GDP) |>
  group_by(Country) |>
  summarise(N_GDP_missing = sum(GDP_missing))

```
