---
title: "In_class_assignment_2"
author: "Shukyee Chan"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Codebook lookup

Look up the codebook, answer the following questions:

What indicators regarding the quality of education are available in the V-Dem datasets?

Answer: 9.1.1 Education 15+ (E) (e_peaveduc) and 9.1.2 Educational inequality, Gini (E) (e_peedgini)

1.  What are the data's coverage (i.e., for which countries and years do we have data?)

    Answer: Years: 1850-2010; Countries: total 202 countries (pp. 40-41)

2.  What are their sources? Provide the link to least 1 source.

    Answer:

    van Leeuwen, B., van Leeuwen-Li, J. & Földvári, P. (2011), 'Regional human capital in Republican and New China: Its spread, quality and effects on economic growth', MPRA Paper 43582, University Library of Munich, Germany .

    URL: <https://mpra.ub.uni-muenchen.de/43582/> <https://ideas.repec.org/p/pra/mprapa/43582.html>

## 2. **Subset by columns**

Create a dataset containing only the country-year identifiers and indicators of education quality.

```{r}
library(tidyverse)
```
