---
title: "In-class assignment 2"
author: "Shukyee Chan"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 1. Codebook lookup

Look up the codebook, answer the following questions:

What indicators regarding the quality of education are available in the V-Dem datasets?

Answer: 9.1.1 Education 15+ (E) (e_peaveduc) and 9.1.2 Educational inequality, Gini (E) (e_peedgini)

1.  What are the data's coverage (i.e., for which countries and years do we have data?)

    Answer: Years: 1850-2010; Countries: total 202 countries (pp. 40-41)

2.  What are their sources? Provide the link to least 1 source.

    Answer:

    van Leeuwen, B., van Leeuwen-Li, J. & Földvári, P. (2011), 'Regional human capital in Republican and New China: Its spread, quality and effects on economic growth', MPRA Paper 43582, University Library of Munich, Germany .

    URL: <https://mpra.ub.uni-muenchen.de/43582/> <https://ideas.repec.org/p/pra/mprapa/43582.html>

## 2. **Subset by columns**

Create a dataset containing only the country-year identifiers and indicators of education quality.

```{r}
library(tidyverse)
```

```{r}
d <- read.csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

```{r}
d |> select(country_name, country_id, year, e_peaveduc, e_peedgini) |> distinct()
```

```{r}
d_edu <- d |> select(country_name, country_id, year, e_peaveduc, e_peedgini)
```

Rename the columns of education quality to make them informative.

```{r}
d_edu <- d_edu |>
  rename("Country" = "country_name", "ID" = "country_id", 
         "Year" = "year", "Education_15" = "e_peaveduc",
         "Education_inequality" = "e_peedgini")
```

## 3. **Subset by rows**

List 5 countries-years that have the highest education level among its population.

```{r}
d_edu |>
  slice_max(order_by = Education_15, n = 5, with_ties = FALSE)
```

List 5 countries-years that suffer from the most severe inequality in education.

```{r}
d_edu |>
  slice_min(order_by = Education_inequality, n = 5)
```

## 4. Summarize the data

Check data availability: For which countries and years are the indicators of education quality available?

```{r}
d_edu |>
  mutate(Education_15_missing = as.numeric(is.na(Education_15)), .after = Education_15)|>
  group_by(Country)|>
  summarise(N_Education_15_missing = sum(Education_15_missing))
```

```{r}
d_edu |>
  mutate(Education_inequality_missing = as.numeric(is.na(Education_inequality)), .after = Education_inequality) |>
  group_by(Country) |>
  summarise(N_Education_inequality_missing = sum(Education_inequality_missing))
```

Average level of education quality from 1984 to 2022

```{r}
d_edu |>
  group_by(Country) |>
  summarise(Education_15_average = mean(Education_15, na.rm = TRUE),
            Education_inequality_average = mean(Education_inequality, na.rm = TRUE))
```

Change of education quality from 1984 to 2022

```{r}
d_edu |>
  filter(Year >= 1984 & Year <= 2010) |>
  group_by(Country) |>
  arrange(Year) |>
  summarise(Education_15_change_1984_2010 = (last(Education_15) - first(Education_15) / first(Education_15)), Education_inequality_change_1984_2010=  (last(Education_inequality) - first(Education_inequality)) / first(Education_inequality)) |>
  ungroup() |>
  arrange(Country)
```

Examine the data and briefly discuss: Which countries perform the best and the worst in terms of education quality in the past four decades?

Answer: The best is Germany and the worst is Botswana.
