---
title: "Lecture7_note_data_visualization"
author: "Shukyee Chan"
date: "2023-10-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Data Preparation

```{r}
library(tidyverse)

d_full <- bind_rows(
  read_csv("_DataPublic_/vdem/1789_1827/vdem_1789_1827_external.csv"),
  read_csv("_DataPublic_/vdem/1867_1905/vdem_1867_1905_external.csv"),
  read_csv("_DataPublic_/vdem/1906_1944/vdem_1906_1944_external.csv"),
  read_csv("_DataPublic_/vdem/1945_1983/vdem_1945_1983_external.csv"),
  read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
)
```

## Select Indicators of Interest

```{r}
d <- d_full |>
  select(
    country_text_id, year, 
    e_regiongeo, e_pelifeex, e_gdppc, 
    e_mipopula, e_wb_pop, e_peinfmor,
    e_boix_regime, e_lexical_index, e_p_polity) |>
  rename("region" = "e_regiongeo", 
         "life_expectancy" = "e_pelifeex", 
         "gdppc" = "e_gdppc",
         "population_ClioInfra" = "e_mipopula", 
         "population_WorldBank" = "e_wb_pop",
         "infant_mortality" = "e_peinfmor",
         "democracy_binary" = "e_boix_regime",
         "democracy_lexical" = "e_lexical_index",
         "democracy_polity5" = "e_p_polity") |>
  filter(year >= 1800)

saveRDS(d, "Lecture_07/data/wealth_and_health.rds")
```

```{r}
summary(d)
```

## Multiple Population Data Sources

```{r}
# Check years that are available in both datasets
d_pop_overlap <- d |> select(country_text_id, year, starts_with("population_")) |> 
  drop_na()
print(d_pop_overlap, n = 3)
```

```{r}
unique(d_pop_overlap$year)
```

```{r}
cor(d_pop_overlap$population_ClioInfra, d_pop_overlap$population_WorldBank)
```

## Set a Rule to Merge the Two Population Columns

-   Different units: Divide `population_WorldBank` by 1000 (so that the unit of population is "in thousands"

-   Different coverage but almost perfect correlation

    -   For years that only one dataset has coverage, take the value from the dataset that has available data points.

    -   For years that both datasets have coverage, take their `mean`.

    -   \--\> taking the `mean` and allow `na.rm = TRUE`.

## Merge the Two Population Columns

```{r}
# STEP 1: "Harmonize" the units
d <- d |> mutate(population_WorldBank = population_WorldBank / 1000)

# STEP 2 Method 1: Slower but use only tidyverse functionality
# [Slow! Not recommended!]
d <- d |> rowwise() |>
  mutate(population = mean(c_across(c("population_ClioInfra", "population_WorldBank")), 
                           na.rm = TRUE), .after = population_WorldBank) |>
  ungroup()

# STEP 2 Method 2: Faster but use a non-tidyverse function rowMeans() 
# and create a temporary vector tmp_population, which I remove after use with rm()
# [Faster !Recommended!]
tmp_population <- d |> select(population_ClioInfra, population_WorldBank) |> rowMeans(na.rm = TRUE)
d <- d |> mutate(population = !!(tmp_population), .after = population_WorldBank)

rm(tmp_population)

# Remove the columns we no longer need
d <- d |> select(-population_ClioInfra, -population_WorldBank)
```

## Sanity Check

```{r}
summary(d %>% select(-country_text_id, -year, -region))
```

## Check Data Availability

```{r}
check_data_available <- d |> 
  mutate(Available = (!is.na(life_expectancy) & !is.na(gdppc) & !is.na(population))) 
# Check number of missing values by country-year
table(check_data_available$Available, useNA = "always")
```

```{r}
check_data_available |> print(n = 3)
```

```{r}
check_data_available_wide <- check_data_available |>
  select(country_text_id, year, Available) |>
  pivot_wider(names_from = "country_text_id", values_from = "Available", 
              names_prefix = "c_") |>
  arrange(year)

check_data_available_wide |> print(n = 3)
```

```{r}
# Check, for each year, the availability of each column
check_data_available_by_column <- d |>
  group_by(year) |>
  summarise(
    life_expectancy = sum(is.na(life_expectancy)),
    gdppc = sum(is.na(gdppc)),
    population = sum(is.na(population))
  )
  # summarise_at(vars(life_expectancy, gdppc, population), ~sum(!is.na(.))) 
  # above is an alternative way to write the summarise() step

check_data_available_by_column |> print(n = 3)
```

## Save Cleaned Data

```{r}
dir.create("Lecture_07/data")
```

```{r}
saveRDS(d, "Lecture_07/data/wealth_and_health.rds")
```

# 2. Data Viz Basics

## Simplest Possible Visualization

```{r}
d |> 
  ggplot(aes(x = gdppc, y = life_expectancy)) + 
  geom_point() +
  labs(x = "GDP per capita", y = "Life Expectancy", 
       title = "Wealth and Health in the World (1800-2019)",
       caption = "By Haohan Chen. Data source: V-Dem v.13")
```

```{r}
# Store in R environment (temporary)
p_all <- d |> 
  ggplot(aes(x = gdppc, y = life_expectancy)) + 
  geom_point() +
  labs(x = "GDP per capita", y = "Life Expectancy", 
       title = "Wealth and Health in the World (1800-2019)",
       caption = "By Haohan Chen. Data source: V-Dem v.13")
# Save plot as a .rds file in your folder
dir.create("Lecture_07/1_data_visualization_1/figures")
saveRDS(p_all, "Lecture_07/1_data_visualization_1/figures/welath_and_health_all.rds")
# Save plot as a PDF file in your folder
ggsave(filename = "Lecture_07/1_data_visualization_1/figures/welath_and_health_all.pdf", 
       plot = p_all, width = 9, height = 4)
```

## Set Themes: `theme_bw`

```{r}
p_all + theme_bw()
```

## Set Themes: `theme_classic`

```{r}
p_all + theme_classic()
```

## Set Themes: `theme_minimal`

```{r}
p_all + theme_minimal()
```

## Other Fancy Themes: The Economist

```{r}
# install.packages("ggthemes") # install the package upon your first use.
# Take a look at the package's website: https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/
library(ggthemes)
p_all + theme_economist()
```

## Other Fancy Themes: The WSJ

```{r}
p_all + theme_wsj(base_size = 6)
```
